---
title: "Operational_Carbon"
author: "Cassandra Etter"
date: '2023-12-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Operational Carbon Projections 

This is the third tranche of work which builds on the indicators RMkdwn in this project. It takes Building stock estimates to 2050 and adds operational carbon estimates. Please refer to first mkdown for earlier work. 

Overview: Data for GlobalABC Project on construction and building indicators and projections
This is the repository containing the source code and documentation for the Global ABC baseline assessment and indicators and projections for gap analysis. 
The point of the project is to 
1) gain a good understanding of the context 
2) identify challenges and opportunities and 
3) measure the objective achievement gap. 

The project was carried out by UNOPS.

##Import library
```{r}
library(data.table)
library(dplyr)
library(countrycode)
library(readxl)
library(tidyr)
library(googlesheets4)
library(tidyverse)
library(wbstats)
library(WDI)
```

##Load Dataset --Clean Cooking Access % 

Pull WHO dataset on primary reliance on clean fuels and technology for cooking which has figures for 1990 - 2021 for 191 countries at the urban/rural/ total % levels. 
Source: https://www.who.int/data/gho/data/indicators/indicator-details/GHO/gho-phe-primary-reliance-on-clean-fuels-and-technologies-proportion 
```{r }
cooking <- read_csv("Data/cooking.csv") 

##create a growth rate function for use. 
cooking <- cooking %>%
        select(SpatialDimValueCode, Period, Dim1, FactValueNumeric) %>%
        arrange(SpatialDimValueCode, Dim1, Period) %>%
        rename("iso3c" = SpatialDimValueCode) 
```

##Create average growth rate for clean cooking access 

To create a prediction to 2050 for this data we will take the growth rate for each year between 1990 - 2021 and then find the average growth rate over that period for urban, rural, and total %. 

```{r}
cooking_total <- cooking %>%
        filter(Dim1 == "Total") %>%
        mutate(growth_rate = (FactValueNumeric/lag(FactValueNumeric)) / 100) %>%
        group_by(iso3c) %>%
        summarise(average_growth = mean(growth_rate[!is.infinite(growth_rate)], na.rm = TRUE))

cooking_rural <- cooking %>%
        filter(Dim1 == "Rural") %>%
        mutate(growth_rate = (FactValueNumeric/lag(FactValueNumeric)) / 100 ) %>%
        group_by(iso3c) %>%
        summarise(average_growth = mean (growth_rate[!is.infinite(growth_rate)], na.rm = TRUE)) 
        
cooking_urban <- cooking %>%
        filter(Dim1 == "Urban") %>%
        mutate(growth_rate = (FactValueNumeric/lag(FactValueNumeric)) /100 ) %>%
        group_by(iso3c) %>%
        summarise(average_growth = mean(growth_rate[!is.infinite(growth_rate)], na.rm = TRUE))

cooking_total_merged <- cooking %>%
        filter(Period == 2020 & Dim1 == "Total" ) %>%
        left_join(cooking_total, by = "iso3c")

cooking_rural_merged <- cooking %>%
        filter(Period == 2020 & Dim1 == "Rural") %>%
        left_join(cooking_rural, by = "iso3c")

cooking_urban_merged <- cooking %>%
        filter(Period == 2020 & Dim1 == "Urban") %>%
        left_join(cooking_urban, by = "iso3c")
```

##Predict to 2050 using average growth rate 
```{r}
years_to_predict <- seq(2020, 2050, by = 1)

cooking_rural_2050 <- cooking_rural_merged %>%
        crossing(year = years_to_predict) %>%
        mutate(Access = pmin( FactValueNumeric * (1 + average_growth) ^ (year - 2020), 100)) %>%
        rename(Rural_Access = Access) %>%
        select(iso3c, year, Rural_Access)

cooking_urban_2050 <- cooking_urban_merged %>%
        crossing(year = years_to_predict) %>%
        mutate(Access = pmin(FactValueNumeric * (1 + average_growth) ^ (year - 2020), 100)) %>%
        rename(Urban_Access = Access) %>%
        select(iso3c, year, Urban_Access)

cooking_total_2050 <- cooking_total_merged %>%
        crossing(year = years_to_predict) %>%
        mutate(Access = pmin(FactValueNumeric * (1 + average_growth) ^ (year - 2020), 100)) %>%
        rename(Total_Access = Access) %>%
        select(iso3c, year, Total_Access)
###merge datasets back together 
Access_Percent <- cooking_total_2050 %>%
        left_join(cooking_urban_2050, by = c("iso3c", "year")) %>%
        left_join(cooking_rural_2050, by = c("iso3c", "year"))
```

Access_Percent is the final dataset used in the projections. 

### Annual Building Stock Operational Energy Consumption (KWh)

Starting point for this analysis is IEA's estimates for 2017 on Building Operational Emissions for 138 countries. This data was pulled from a data file from (Guo 2020) "Global comparison of building energy use data within the context of climate change". 
```{r}
Global_Buildings_Data <- read_excel("Data/Global_Buildings_Data.xlsx") %>%
        rename(Country = 1) %>%
        mutate(iso3c = countrycode(Country, "country.name", "iso3c")) %>%
        subset(Country != "Kosovo" & Country != "World") 
```
We then need to create a regional identifier for each country because we will have many 50+ missing countries as the IEA dataset only has 138 and we look at 196 countries 
```{r}
countries <- unique(Global_Buildings_Data$iso3c)
regions <- countrycode(countries, "iso3c", "region")
result <- data.frame(Country = countries, Region = regions)
```
formatting data and adding a new column to align units 

```{r}
Final_Energy <- Global_Buildings_Data %>%
        select(Country, iso3c, `FEC-Total`, FEC_Residential, `FEC_Commerce and public services`, FEC_Buildings) %>%
        subset(Country != "Unit") %>%
        pivot_longer(c("FEC-Total", "FEC_Residential", "FEC_Commerce and public services", "FEC_Buildings"), names_to = "Final Energy Consumption", values_to = "ktoe") %>%
        mutate(tCO2e = as.numeric(ktoe) * 100) %>%
        left_join(result, by = c("iso3c" = "Country")) %>%
        filter(`Final Energy Consumption` == "FEC_Buildings")


Total_Carbon <- Global_Buildings_Data %>%
        select(Country, iso3c, Total_CarbonE, Residential_CarbonE, `Commerce and public services_CarbonE`, Building_Carbon) %>%
        subset(Country != "Unit") %>%
        pivot_longer(c("Total_CarbonE", "Residential_CarbonE", "Commerce and public services_CarbonE", "Building_Carbon"), names_to = "Carbon Emissions", values_to = "MtCO2e") %>%
        mutate(Year = 2017, tCO2e = as.numeric(MtCO2e) * 1e6)%>%
        left_join(result, by = c("iso3c" = "Country")) %>%
        filter(`Carbon Emissions` == "Building_Carbon")
       


Direct_Carbon_Emissions <- Global_Buildings_Data %>%
        select(Country, iso3c, Residential_DirectCE, `Commerce and public services_DirectCE`, Building_DirectCE) %>%
        subset(Country != "Unit") %>%
        pivot_longer(c("Residential_DirectCE", "Commerce and public services_DirectCE", "Building_DirectCE"), names_to = "Direct Carbon Emissions", values_to = "MtCO2e") %>%
        mutate(Year = 2017, tCO2e = as.numeric(MtCO2e) * 1e6) %>%
        left_join(result, by = c("iso3c" = "Country")) %>%
        filter(`Direct Carbon Emissions` == "Building_DirectCE")
```
###Projecting Operational Carbon to 2050 ####

We now have data for 2017, but need to project out to 2050. We decide here to make the 2017 data the baseyear (2020), though recognize this does not match. It is the latest data we could find. Securing the IEA data for 2020 would help improve this estimate. 

```{r}
###approach to 2050 taking operational emissions to be constant. building floor space to change of total emissions 

bs_2050 <- read_csv("Data/BS_2050.csv") %>%
        select(`Country Name`, `iso3c`, year, `total_building_stock_m2` ) 

final_bs <- bs_2050 %>% filter(year == 2020) %>%
  left_join(Final_Energy %>% select(iso3c, `Building Energy Consumption (toe)` = tCO2e), by = "iso3c") %>%
  left_join(Total_Carbon %>% select(iso3c, `Building Carbon Emissions (tCO2e)` = tCO2e), by = "iso3c") %>%
  left_join(Direct_Carbon_Emissions %>% select(iso3c, `Direct Building Emissions (tCO2e)` = tCO2e), by = "iso3c") %>%
  mutate(`BEC (toe/m2)` = `Building Energy Consumption (toe)`/ `total_building_stock_m2`, 
         `BCE (tCO2e/m2)` = `Building Carbon Emissions (tCO2e)` / `total_building_stock_m2`, 
         `DBE (tCO2e/m2)` = `Direct Building Emissions (tCO2e)` / `total_building_stock_m2`) %>%
  select(iso3c, `BEC (toe/m2)`, `BCE (tCO2e/m2)`, `DBE (tCO2e/m2)`) %>%
  left_join(bs_2050, by = "iso3c") %>%
  mutate(`Building Energy Consumption (toe)` = `BEC (toe/m2)` * total_building_stock_m2,
         `Building Carbon Emissions (tCO2e)` = `BCE (tCO2e/m2)` * total_building_stock_m2,
         `Direct Building Emissions (tCO2e)` = `DBE (tCO2e/m2)` * total_building_stock_m2) %>%
  arrange(iso3c, year) %>%
  select(`Country Name`, iso3c, year, everything())



```
We now have a dataset with building carbon emissions, as well as the sub-components of final energy consumption and direct building emissions. 

UNOPS here decide to take full building carbon emissions and move forward with only this estimate, as it is the broaded and encompasses the other two components of operational emissions. 

We are, however, lacking 50+ countries which need to be added manually using a rate of change information for operational emissions. the following regional estimates are applied to the regions [STOPPING POINT FOR 2023]

```{r}
###this provides a 2017 figure for Final Energy Consumption, Total Carbon Emissions, and Direct Carbon Emissions. To find 2050 estimate need to secure how they are projected to decrease or increase to 2050. 

#Need rate of change information for Total Carbon (Total, Residential, Commercial, Buildings)

#Europe and Central Asia
#Middle East and North Africa
#Sub saharan Africa
#LATAM and Caribbean
#East Asia and Pacific
#South Asia
#North America
```

Combine that dataset with our cooking estimates earlier for the operational carbon sheet to be imported to google sheets. 

```{r}
access <- read_csv("Data/Access_2050.csv") %>%
        rename("iso3c" = "Country", "year" = "Year")

operational_2050_combined <- access %>% 
        left_join(final_bs, by = c("iso3c", "year"))

write_csv(operational_2050_combined, "Operational_Emissions_Estimates.csv") 
```