---
title: "Adaptation"
author: "Cassandra Etter"
date: "2024-03-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Adaptation Indicators

This is the fifth tranche of work which builds on the indicators related to adaptation.

Overview: Data for GlobalABC Project on construction and building indicators and projections
This is the repository containing the source code and documentation for the Global ABC baseline assessment and indicators and projections for gap analysis. 
The point of the project is to 
1) gain a good understanding of the context 
2) identify challenges and opportunities and 
3) measure the objective achievement gap. 

The project was carried out by UNOPS.

## Import Libraries 
```{r}

library(tidyverse)
library(dplyr)
library(wbstats)
library(readxl)
library(countrycode)
library(googlesheets4)
library(forecast)
```

##2.6 - Population Affected by National Disaster in last three years (% of population)####

We use EMDAT dat which tracks naturaldisasters and calculate for 2020- 2023 to find the number of people affected. we then turn this into a % of the total population. 

To do this we will aggregate all natural disasters: Biological (ex. locust infestation) hydrological (ex. flood) meteorological (ex. storm) geophysical (Volcano or earthquake), or climatological (ex. wildfire). 

We also aggregate number affected which is total deaths, injuries, and displaced people. 
```{r}
nat_disaster <- read_xlsx("~/Constructionprojections/Data/vulnerability_emdat.xlsx") 

summary_nat <- nat_disaster%>%
        group_by(Country, ISO) %>%
        summarize(
                total_events = n(),
                total_affected = sum(`Total Affected`)
        )
print(nat_disaster)

pop <- wb_search("TOTAL POPULATION")
pop <- wb_data("SP.POP.TOTL", start_date = 2022, end_date = 2022)%>%
        select(Country = country, iso3c, pop = SP.POP.TOTL)

nat_disaster <- pop %>%
        left_join(summary_nat, by = "Country") %>%
        mutate(Percent = `total_affected` / pop)

```

Decision now to move to inform risk index metric as our data doesn't appear to be correct. 

```{r}
natural <- read_xlsx("~/ConstructionProjections/Data/NaturalDisaster_RiskIndex.xlsx") 
natural <- head(natural, -2)

Adaptation_Final <-natural %>%
        select(Country = COUNTRY, iso3c = ISO3, nat_disaster_2.6 = `Natural Disasters % of total pop`)
write_csv(Adaptation_Final, "~/ConstructionProjections/Data/Adaptation_FINAL.csv")
```
##Domestic Annual Extractionof non-metallic minerals 5.4
Using UNEP resource panel global material flows database for estimates of projected resource extraction. 
```{r}
extraction <- read_csv("~/Downloads/mfa4_export (3).csv")
extraction <- extraction %>%
        select(-`Flow name`, - `Flow code`, - `Flow unit`) %>%
        pivot_longer(!c(Country, Category), names_to = "year", values_to = "extraction" )


extraction$year <- as.numeric(extraction$year)

# Filter data for years up to 2020 (or any last historical year)
historical_data <- extraction %>%
  filter(year <= 2020)

# Group by Country and Category, and fit linear regression model to historical data
linear_models <- historical_data %>%
  group_by(Country, Category) %>%
  do(model = lm(extraction ~ year, data = .))

# Create a data frame to store projected values
projection_data <- data.frame(
  Country = character(),
  Category = character(),
  Year = numeric(),
  Extraction = numeric()
)

# For each country and category, predict extraction values for years 2021 to 2050
for (i in seq_along(linear_models$model)) {
  country <- linear_models$Country[i]
  category <- linear_models$Category[i]
  model <- linear_models$model[[i]]
  
  # Create data frame with years 2021 to 2050
  future_years <- data.frame(year = 2021:2050)
  
  # Predict extraction values for future years using the linear regression model
  # Ensure column names match the original data used to fit the model
  names(future_years) <- c("year")
  
  # Predict extraction values for future years using the linear regression model
  projected_values <- predict(model, newdata = future_years)
  
  # Append projected values to projection_data data frame
  projection_data <- bind_rows(projection_data, data.frame(
    Country = country,
    Category = category,
    Year = future_years$year,
    Extraction = projected_values
  ))
}

# Print the projected data
print(projection_data)
```

After creating a linear projection, add in Iso3c and clean up for export. 

```{r}
Adaptation_Projection_Final <- projection_data%>%
        mutate(iso3c = countrycode(Country, "country.name", "iso3c")) %>%
        select(Country, iso3c, Year, extraction_5.4 = Extraction) 
        

write_csv(Adaptation_Projection_Final, "~/ConstructionProjections/Data/Adaptation_Projection_FINAL.csv")
```

