---
title: "Embodied_Carbon"
author: "Cassandra Etter"
date: '2023-12-06'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Embodied Carbon Projections 

This is the second tranche of work which builds on the indicators RMkdwn in this project. It takes Building stock estimates to 2050 and adds embodied carbon estimates. Please refer to first mkdown for earlier work. 

Overview: Data for GlobalABC Project on construction and building indicators and projections
This is the repository containing the source code and documentation for the Global ABC baseline assessment and indicators and projections for gap analysis. 
The point of the project is to 
1) gain a good understanding of the context 
2) identify challenges and opportunities and 
3) measure the objective achievement gap. 

The project was carried out by UNOPS.

##import library
```{r}
library(data.table)
library(dplyr)
library(countrycode)
library(readxl)
library(tidyr)
library(googlesheets4)
library(tidyverse)
library(stringr)
library(bit64)
```

## 4.1 Annual building stock embodied carbon emissions-- Existing Buildings

To find embodied carbon estimates by country we use EXIOBASE3, a multi-regional input output database with access to 47 countries and 5 regions. 
```{r}
### importing MRIO into working environment 

D_cba <- fread("~/Downloads/IOT_2022_ixi/satellite/D_cba.txt")
D_cba_impact <- fread("~/Downloads/IOT_2022_ixi/impacts/D_cba.txt")

M_impact <- fread("~/Downloads/IOT_2022_ixi/impacts/M.txt")

exio3_iso2 <- fread("~/Downloads/IOT_2022_ixi/impacts/D_cba.txt", 
                    nrows = 1, header = FALSE) %>% unlist %>% tail(-1)

exio3_sectors <- fread("~/Downloads/IOT_2022_ixi/impacts/D_cba.txt", 
                    nrows = 2, header = FALSE)[2,] %>% unlist %>% tail(-1)


D_cba_impact[region == "GHG emissions (GWP100) | Problem oriented approach: baseline (CML, 2001) | GWP100 (IPCC, 2007)"]

exio_ghg <- data.table(
        iso3 = exio3_iso2 %>% countrycode("iso2c", "iso3c"),
        sector = exio3_sectors,
        ghg_int = M_impact[region == "GHG emissions (GWP100) | Problem oriented approach: baseline (CML, 2001) | GWP100 (IPCC, 2007)"] %>% 
                unlist %>% tail(-1) %>% as.numeric,
        ghg_tot = D_cba_impact[region == "GHG emissions (GWP100) | Problem oriented approach: baseline (CML, 2001) | GWP100 (IPCC, 2007)"] %>% 
                unlist %>% tail(-1) %>% as.numeric) 

exio_ghg <- exio_ghg %>%
        mutate(Country = countrycode(iso3, "iso3c", "country.name"))
```

###Pulling relevant products from MRIO 

We now pull subset for Construction, Manufacture of cement, lime and plaster, Manufacture of basic iron and steel and of ferro-alloys and first products thereof Aluminium production. This is based on UNOPS recommendation for these products accounting for most embodied emissions in buildings. 
```{r}
list_substet <-c("Manufacture of cement, lime and plaster", "Aluminium production" ,"Construction (45)","Manufacture of basic iron and steel and of ferro-alloys and first products thereof")
subset_exio_ghg <- exio_ghg %>% subset(sector %in% list_substet)
```

We use this to create variable for construction emissions intensity and building emissions intensity 
```{r}
Building_intensity <- subset_exio_ghg %>%
        subset(sector %in% c("Manufacture of cement, lime and plaster", "Aluminium production","Manufacture of basic iron and steel and of ferro-alloys and first products thereof")) %>%
        group_by(iso3) %>%
        summarise(
                Building_GHG_Intensity = sum(ghg_int),
                Building_GHG_Total = sum(ghg_tot)
        ) %>%
        rename(iso3c = iso3)
Construction_intensity <-subset_exio_ghg %>%
        subset( sector %in% "Construction (45)") %>%
        rename(
                Construction_GHG_Intensity = ghg_int,
                Construction_GHG_Total = ghg_tot,
                iso3c = iso3
        ) %>%
  select(-Country)
```

###Import ICP Data on Costs

We now use the World Bank's International Comparison Programme data. This data collects comparative price data and detailed GDP expenditures to produce a PPP. This will allow us to take the import/ export data from above and allow for comparison across different countries. All goods are benchmarked to 2017.


For missing country data, we apply the listed regional assumptions. We then compare all countries against one country for the comparison, and select Germany as the country selected does not matter for the analysis. Regional assumptions and their comparison cost are below: 

East Asia and Pacific	91.774	967.82
Europe and Central Asia	120.756	1,273.45
Latin America and Caribbean	88.581	934.15
Middle East and North Africa	60.452	637.51
North America	249.646	2,632.69
Sub Saharan Africa	67.229	708.98
South Asia 	42.897	452.38
World (Other) 	100	1,054.57


```{r}
construction_costs <- read_csv("~/ConstructionProjections/Data/construction_costs.csv") 
BS_2050 <- read.csv("~/ConstructionProjections/Data/BS_2050.csv")

Costs_2 <- BS_2050 %>%
        left_join(construction_costs, by = c("Country","iso3c")) %>%
        mutate(Total_Cost = total_building_stock_m2 * `Construction Comparative`) 

```

##Estimate Embodied Emissions of buildings and construction 

Bring in building stock information from first Markdown and add in construction intensity data add building intensity data to this dataset 

```{r}
Embodied_Carbon <- Costs_2 %>%
        left_join(Construction_intensity, by =  "iso3c") %>%
        left_join(Building_intensity, by = "iso3c") %>%
        select(c(`Country`, iso3c, year, total_building_stock_m2, `Construction Comparative`, Total_Cost, 
                 Construction_GHG_Intensity, Building_GHG_Intensity, ))
```
From here there are a number of missing data that need to be added on embodied construction and building intensity. Take regional estimates and apply them to countries in that region. The regional figures provided are:  

East Asia and Pacific	367489.6482	14047106.88
Europe and Central Asia	429531.0633	4688629.699
Latin America and Caribbean	487166.6121	13408428.28
Middle East and North Africa	938445.1	15011385
North America	284101.1124	8734324.867
Sub Saharan Africa	899792.8	62727133
South Asia      576581.4669 14274132.53

```{r}
Embodied_Carbon_wRegion <- Embodied_Carbon %>%
          mutate(region = countrycode(sourcevar = iso3c, origin = "iso3c", destination = "region"))

regional_averages <- data.frame(
        region = c("East Asia & Pacific", "Europe & Central Asia", "Latin America & Caribbean", "Middle East & North Africa", "North America", "Sub-Saharan Africa", "South Asia"), 
        avg_construction_intensity = c(367489.6482, 429531.0633, 487166.6121, 938445.1, 8734324.867, 899792.8, 576581.4669), 
        avg_building_intensity = c(14047106.88, 4688629.699, 13408428.28, 15011385, 8734324.867, 62727133, 14274132.53))

Embodied_Carbon_v2 <- Embodied_Carbon_wRegion %>%
  left_join(regional_averages, by = "region") %>%
  mutate(
    Construction_Intensity = ifelse(is.na(Construction_GHG_Intensity), avg_construction_intensity, Construction_GHG_Intensity),
    Building_Intensity = ifelse(is.na(Building_GHG_Intensity), avg_building_intensity, Building_GHG_Intensity)
  ) %>%
  select(-avg_construction_intensity, -avg_building_intensity)
```

Combine together the various data and put in the same unit of measure for a Ttoal Embodied Carbon Buildings (TCO2e) and Total Embodied Carbon Construction (TCO2e)

```{r}
Embodied_Carbon_v2 <- Embodied_Carbon_v2 %>%
        mutate(Embodied_Carbon_Buildings = `Total_Cost`/ 1000000 * `Building_Intensity` / 1000) %>%
        mutate(Embodied_Carbon_Construction = `Total_Cost` /1000000 * `Construction_Intensity` / 1000) %>%
        select(Country, iso3c, year, total_building_stock_m2, `Construction Comparative`, Total_Cost, Construction_Intensity, Building_Intensity, Embodied_Carbon_Buildings, Embodied_Carbon_Construction)

```

### Average annual embodied carbon per m2 (4.2) -- 

Average annual embodied carbon is just an easier way of seeing emissions intensity per m2. Its a constant. 
```{r}

Embodied_FINAL <- Embodied_Carbon_v2 %>%
  mutate(average_per_m2 = Embodied_Carbon_Buildings / total_building_stock_m2) %>%
  rename(Embodied_Building_4.1 = Embodied_Carbon_Buildings,
         Average_4.2 = average_per_m2) %>%
  select(`Country`, iso3c, year, `Embodied_Building_4.1`, Average_4.2)

```

##Urban Land Area -- Spatial and Urban Development Embodied Carbon (2.1)
Now we turn to urban and spatial development area and the urban land area. To calculate this we take the World Bank urban land area data and apply the world urbanization prospectus change rate for urban to each country. 
Dataset 1: World bank -- urban land area (sq.km) at 5 year intervals. We pull most recent observation from 2015. 

```{r}
urban_land <- read.csv("~/ConstructionProjections/Data/NationalTotalUrbanArea_SSP2_2000-2100_v1.csv") %>%
  pivot_longer(cols = -c(`ISO3v10`, Country), 
               names_to = "year",
               values_to = "km2") %>%
  mutate(year = str_remove(year, "UrbAmt")) %>%
  rename(iso3c = ISO3v10) %>%
  filter(as.numeric(year) <= 2050) 

urban_land$km2 <- as.numeric(urban_land$km2)
urban_land$year <- as.numeric(urban_land$year)

urban_annual <- CJ(year = 2000:2050, Country = unique(urban_land$Country)) %>%
   mutate(base_year = ((year - 2000) %/% 10) * 10 + 2000) %>%
  left_join(urban_land %>% select(Country, base_year = year, base_km2 = km2) %>%
              mutate(annual_increase = (lead(base_km2) - base_km2) / 10), 
            by = c("Country", "base_year")) %>%
  # Special case for 2050
  mutate(km2 = ifelse(year < 2050, base_km2 + annual_increase * (year - base_year), base_km2)) %>%
  arrange(Country, year) %>%
  left_join(urban_land %>% select(Country, iso3c) %>% unique, by = "Country")


urban_annual <- urban_annual%>% arrange(Country, year) %>%
  select(Country, iso3c, year, km2)

Embodied_FINAL <- Embodied_FINAL %>%
  left_join(urban_annual, by = c("Country", "iso3c", "year")) %>%
  rename(`urb_land_2.1` = `km2`)
```

## Urban Population Density -- Spatial and Urban Development Embodied Carbon (2.2)

First we take urban population projections at mid-year by country 1950 - 2050 (thousands) from World Urbanization Prospectus population.un.org. We then combine that with 4.1 to derive the urban density estimate. 

```{r}
urb_pop <- read_excel("~/ConstructionProjections/Data/WUP_UrbanPopulation_Projection.xls") %>%
  pivot_longer(cols = -Country, 
               names_to = "year",
               values_to = "population") %>%
   mutate(iso3c = Country %>% countrycode("country.name", "iso3c")) %>%
  filter(as.numeric(year) >= 2020)

urb_pop$year = as.numeric(urb_pop$year)
  
urb_density <- urb_pop %>%
  left_join(urban_annual, by = c("Country", "iso3c", "year")) %>%
  mutate(density_2.2 = (population * 1000) / km2) 


Embodied_FINAL <- Embodied_FINAL %>%
  left_join(urb_density, by = c("Country", "iso3c", "year")) %>%
  select(-c(population, km2 ))

rm(urb_density, urb_pop, urban_annual, urban_land, subset_exio_ghg, regional_averages, M_impact, exio_ghg, Embodied_Carbon, Embodied_Carbon_v2, Embodied_Carbon_wRegion, D_cba, D_cba_impact, Costs_2, Construction_intensity, construction_costs, Building_intensity, urb_pop)
```

## Carbon emissions of construction materials (5.1, 5.2, 5.3)

Now move to embodied emissions of construction. This data was not clean at all, and exiobase3 did not provide country level emissions intensity or material intensity information to allow for global analysis. The modified approach was derived below after the first approach was too sporatic to move forward with due to lack of data. 

```{r}
# Emission intensity in tCO2 per ton of Non-Metallic Mineral
exio3_F <- data.table(
  iso2 = fread("~/Downloads/IOT_2022_ixi/satellite/F.txt", nrows = 1, header = FALSE)[1, ] %>%
    unlist %>% tail(-1),
  sector = fread("~/Downloads/IOT_2022_ixi/satellite/F.txt", nrows = 2)[1, ] %>% 
    unlist %>% tail(-1),
  non_metallic = fread("~/Downloads/IOT_2022_ixi/satellite/F.txt", skip = 3) %>%
    .[str_detect(V1, "Domestic Extraction Used - Non-Metallic")] %>% select(-1) %>% as.matrix %>% colSums,
  ghg = fread("~/Downloads/IOT_2022_ixi/satellite/F.txt", skip = 3) %>%
    .[V1 %in% c("CO2 - combustion - air", "CH4 - combustion - air", "N2O - combustion - air",
                "CO2 - non combustion - Cement production - air",
                "CO2 - non combustion - Lime production - air")] %>% 
    select(-1) %>% as.matrix %>% colSums
)

# Compute intensity
# - non_metallic is in kt (thousand metric tons)
# - ghg is in kg
# Ensure that intensity is expressed in tCO2 per ton of material

materials_ghg_intensity <- exio3_F[sector %in% c("Quarrying of stone", "Quarrying of sand and clay",
                      "Mining of chemical and fertilizer minerals, production of salt, other mining and quarrying n.e.c."), 
        .(non_metallic = sum(non_metallic) * 1e3)] %>% 
  cbind(
    exio3_F[sector %in% c("Quarrying of stone", "Quarrying of sand and clay",
                          "Mining of chemical and fertilizer minerals, production of salt, other mining and quarrying n.e.c.")] %>%
      .[, .(ghg_extraction = sum(ghg)/1e3)]
  ) %>%
  cbind(
    exio3_F[sector %in% c("Manufacture of bricks, tiles and construction products, in baked clay",
                          "Manufacture of cement, lime and plaster", "Re-processing of ash into clinker",
                          "Manufacture of other non-metallic mineral products n.e.c.")] %>%
      .[, .(ghg_prodution = sum(ghg)/1e3)]
  ) %>%
  mutate(ghg_int_extraction = ghg_extraction / non_metallic,
         ghg_int_prodution  = ghg_prodution / non_metallic)

ghg_int_extraction <- materials_ghg_intensity$ghg_int_extraction
ghg_int_production <- materials_ghg_intensity$ghg_int_prodution

```
From here it is clear that ghg intensity of extraction and production are only available at global level. use the following as obtained from Exiobase 3 above, in tCO2 per ton of non-metallic minerals

ghg_int_extraction <- 0.004537613
ghg_int_production <- 0.124635

###Add UNEP domestic material input information 

Domestic material input is domestic extraction + physical imports (direct, territorial)
```{r}
ghg_int_extraction <- 0.004537613
ghg_int_production <- 0.124635
  
Embodied_extraction <-  BS_2050 %>% 
  # Will apply total building stock evolution over time to non-metallic minerals
  left_join(BS_2050 %>% filter(year == 2020) %>%
              select(Country, total_building_stock_m2_2022 = total_building_stock_m2), 
            by = "Country") %>%
  # Join UNEP data
  left_join(
    fread("~/Downloads/mfa4_export.csv")[
      V3 %in% c("Domestic Extraction", "Imports") &
        V2 == "Non-metallic minerals"
    ] %>%
      # Only keep 2022
      select(Country = V1, Flow = V3, non_metallic = V58) %>%
      dcast(Country ~ Flow, value.var = "non_metallic") %>%
      mutate(`Domestic share` = `Domestic Extraction` / (`Domestic Extraction` + Imports)) %>%
      rename(dom_non_metallic = `Domestic Extraction`,
             imp_non_metallic = Imports),
    by = "Country"
  ) %>%
  # Evolution over time of domestically extracted and imported non-metallic minerals
  mutate(dom_non_metallic = dom_non_metallic * total_building_stock_m2 / total_building_stock_m2_2022,
         imp_non_metallic = imp_non_metallic * total_building_stock_m2 / total_building_stock_m2_2022,
         dom_non_metallic_ghg_extraction = dom_non_metallic * ghg_int_extraction,
         dom_non_metallic_ghg_tot = dom_non_metallic * (ghg_int_extraction + ghg_int_production),
         imp_non_metallic_ghg_extraction = imp_non_metallic * ghg_int_extraction,
         imp_non_metallic_ghg_tot = imp_non_metallic * (ghg_int_extraction + ghg_int_production))

```

This version created serious problems. But keeping it for reference. Developed final method after trial and error for magic. 
###Corrected Version of embodied construction without errors 

```{r}

# Convert integer64 columns to numeric
UNE_data <- UNE_data %>% mutate(across(where(is.integer64), as.numeric))

# Perform calculations with the updated data frame
Embodied_extraction <- BS_2050 %>% 
  # Will apply total building stock evolution over time to non-metallic minerals
  left_join(BS_2050 %>% filter(year == 2022) %>%
              select(Country, total_building_stock_m2_2022 = TBS), 
            by = "Country") %>%
  # Join UNEP data
  left_join(
    UNE_data[
      V3 %in% c("Domestic Extraction", "Imports") &
        V2 == "Non-metallic minerals"
    ] %>%
      # Only keep 2022
      select(Country = V1, Flow = V3, non_metallic = V58) %>%
      dcast(Country ~ Flow, value.var = "non_metallic") %>%
      mutate(`Domestic share` = `Domestic Extraction` / (`Domestic Extraction` + Imports)) %>%
      rename(dom_non_metallic = `Domestic Extraction`,
             imp_non_metallic = Imports),
    by = "Country"
  ) %>%
  # Evolution over time of domestically extracted and imported non-metallic minerals
  mutate(dom_non_metallic = dom_non_metallic * TBS / total_building_stock_m2_2022,
         imp_non_metallic = imp_non_metallic * TBS / total_building_stock_m2_2022,
         dom_non_metallic_ghg_extraction = dom_non_metallic * ghg_int_extraction,
         dom_non_metallic_ghg_tot = dom_non_metallic * (ghg_int_extraction + ghg_int_production),
         imp_non_metallic_ghg_extraction = imp_non_metallic * ghg_int_extraction,
         imp_non_metallic_ghg_tot = imp_non_metallic * (ghg_int_extraction + ghg_int_production))

```

#### Review data and check it for completeness 

```{r}
select_countries <- c("India", "China", "Haiti", "Canada", "Zimbabwe", "United States")

ggplot(Embodied_extraction %>% filter(Country %in% select_countries), aes(x = year, y= dom_non_metallic, group = Country, color = Country)) +
  geom_line()

ggplot(Embodied_extraction %>% filter(Country %in% select_countries), aes(x = year, y = dom_non_metallic_ghg_tot, group = Country, color = Country)) +
  geom_line()
```
Discover if any missing variables. Discover that 19 countries are missing from dataset. Lao, St. Lucia, West Bank, Slovak Republic, Turkiye, United States, St. Vincent and the Grenadines, Vietnam, Yemen 

```{r}
NA_countries <- Embodied_extraction %>%
  filter(is.na(dom_non_metallic_ghg_tot)) %>%
  distinct(Country)
print(NA_countries)
```
### Carbon emissions of imported construction materials (5.2)

imp_non_metallic_ghg_tot = 5.2 
### Ratio of nationally produced and imported construction materials (5.3)

Domestic Share = 5.3

Final data used is here under Embodied_FINAL.csv 

```{r}

Embodied_FINAL <- Embodied_FINAL %>%
  left_join(Embodied_extraction, by = c("Country", "iso3c", "year")) %>%
  select(Country, iso3c, year, urb_land_2.1, density_2.2, Embodied_Building_4.1, Average_4.2, Construction_5.1 = dom_non_metallic_ghg_tot, Const_Import_5.2 = imp_non_metallic_ghg_tot, Ratio_5.3 = `Domestic share`)

write_csv(Embodied_FINAL, "~/ConstructionProjections/Data/Embodied_FINAL.csv")
```

