---
title: "Global ABC Roadmap Development -- Baseline Assessment"
author: "Cassandra Etter"
date: '2023-10-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown for UNOPS Buildings Data 

Data for GlobalABC Project on construction and building indicators and projections
This is the repository containing the source code and documentation for the Global ABC baseline assessment and indicators and projections for gap analysis. 
The point of the project is to 
1) gain a good understanding of the context 
2) identify challenges and opportunities and 
3) measure the objective achievement gap. 

The project was carried out by UNOPS.

###Import Necessary Packages
```{r}
library(readxl)
library(dplyr)
library(tidyverse)
library(devtools)
library(wbstats)
library(WDI)
library(readr)
library(remotes)
#library(wpp2022)
library(countrycode)
library(ggplot2)
library(fabletools)
library(fable)
library(tsibble)
library(tsibbledata)
library(lubridate)
library(devtools)
library(feasts)
library(googlesheets4)
library(data.table)
```

# Building Stock Floor area m2 baseline measurement 

The goal is to build a floor area m2 per country using indicators for gdp*capita, household size, and population. 

For GDP, I use World Bank API and pull GDP per capita, PPP (current international $) for 2020. For future updating, change this figure to the most recent available year. 

##importing gdp per capita, PPP (current international $)
```{r}
gdp_per_capita <- wb_search("gdp.*capita")
gdp_per_capita <- wb_data("NY.GDP.PCAP.PP.CD", start_date = 2020, end_date = 2020)
```

For Population, I use World Bank data for base year 2020. 

##importing population
```{r}
Population <- wb_search("population")
Population <- wb_data(indicator = "SP.POP.TOTL", start_date = 2020, end_date = 2020)
head(Population)


```

To retrieve average household size, I use UNDESA Data in baseyear 2020. As the data is survey based, some of the estimates in the dataset are from earlier years than 2020, in which case the most recent year is selected. 

##importing average household size 
```{r}
undesa_pd_2022_hh_size_composition <- read_excel("~/ConstructionProjections/Data/undesa_pd_2022_hh-size-composition.xlsx", 
                                                 sheet = "HH size and composition 2022", 
                                                 range = "A5:E1064")
average_household <- undesa_2022 <- read_excel("~/Documents/GlobalABC project/undesa_2022.xlsx", 
                                               sheet = "HH size and composition 2022", range = "A5:E1064")
average_household <- as.data.frame(average_household)
average_household <- average_household%>% 
        arrange(desc(`Reference date (dd/mm/yyyy)`)) %>%
        group_by(`ISO Code`) %>%
        filter(row_number()==1) %>%
        drop_na() %>%
        mutate(iso3c = countrycode(`Country or area`, "country.name", "iso3c")) 

head(average_household)
```

This is a very messy dataset, and on review requires average household sizes be added for Botswana, Brunei, Bulgaria, China, French Polynesia, Iran, New Caledonia, Palau, Seychelles, St. Kitts, Trinidad and Tobago. I've checked national statistics websites to add missing variables. As there is also only one observation period for most countries, we are unable to projected trends in these figures. Moving forward, we will keep average household size constant, but assume that population growth will continue to expand number of households for projection purposes. 
##adding missing data to average household size 
```{r}
average_household$`Average household size (number of members)`[average_household$iso3c =="CHN"] <- 2.76 
average_household$`Average household size (number of members)`[average_household$iso3c =="IRN"] <- 3.2 
average_household$`Average household size (number of members)`[average_household$iso3c =="BWA"] <- 3.3 
average_household$`Average household size (number of members)`[average_household$iso3c =="BRN"] <- 5 
average_household$`Average household size (number of members)`[average_household$iso3c =="BGR"] <- 2.2 
average_household$`Average household size (number of members)`[average_household$iso3c =="COK"] <- 4.4 
average_household$`Average household size (number of members)`[average_household$iso3c =="KNA"] <- 3.7 
average_household$`Average household size (number of members)`[average_household$iso3c =="TTO"] <- 3.2 
average_household$`Average household size (number of members)`[average_household$iso3c =="CYM"] <- 2.4 
average_household$`Average household size (number of members)`[average_household$iso3c =="SYC"] <- 3.7 
average_household$`Average household size (number of members)`[average_household$iso3c =="SHN"] <- 2.38 
average_household$`Average household size (number of members)`[average_household$iso3c =="PYF"] <- 2.3 
average_household$`Average household size (number of members)`[average_household$iso3c =="NCL"] <- 3.2 
average_household$`Average household size (number of members)`[average_household$iso3c =="PLW"] <- 2.9 

average_household <- average_household %>%
        drop_na(iso3c) 
average_household$`Average household size (number of members)`= as.numeric(average_household$`Average household size (number of members)`)

head(average_household)
```

## Calculate Average Buildingstock baseline 

With all data imported and cleaned, we can now merge data to estimate the average building stock baseline (m2) as an estimate of residential floor area (m2)
```{r}
merged_data <- average_household %>%
        full_join(Population, by = "iso3c") %>%
        full_join(gdp_per_capita, by = "iso3c") %>%
        select(c(1, 2, 6, 5, 10, 18))

building_stock_m2 <- merged_data %>%
        mutate(average_household_size = as.numeric(`Average household size (number of members)`)) %>%
        mutate(num_households = SP.POP.TOTL / average_household_size) %>%
        mutate(avg_m2_house = NY.GDP.PCAP.PP.CD / 1000) %>%
        mutate(total_building_stock_m2 = num_households * avg_m2_house)
```

##Check Data
To review our estimate, we look at the EU Building Stock Observatory Data Online, which provides estimated building stock information for each EU country. We sub select Germany, Finland and Poland at random for comparison. 
EU Observatory Data for total: DEU listed as 5,395,110,000 m2 for 2020, FIN =  311,270,000 , POL = 1,412,210,000 
EU Observatory Data for residential alone: DEU = 3,566,430,000 FIN is = 205,660,000 POL = 1,009,270,000
*Our estimate:* DEU 2,194,921,435 , FIN = 139,721,906 POL = 474,621,575 


#Building Stock Projections to 2050

We now turn to projecting our building stock out to 2050. Rather than regressing estimates on temporal data for building stock, we assume clear relationship between building stock and population growth, economic growth, and other social factors. As such, we use existing "best in class" prediction data from World Bank and the UNFCCC modelling group responsible for the Shared Socioeconomic Pathways used the AR6 report. This allows us to predict building stock to 2050 while incorporating more precise predictions. 

##calculate projection for population figures using world bank projections out to 2050 ##
```{r}
population_projection <- read_excel("~/Documents/GlobalABC project/WB_Population_Projections.xlsx")

Total_pop_2050 <- population_projection %>%
        filter(`Series Name` == 'Population, total') %>%
        gather(Year, Population, -`Country Name` & 
                       - `Country Code` & - `Series Name` & -`Series Code`) %>%
        mutate(cool_year = str_sub(Year, 1, 4) %>% 
                       str_c(., "-01-01") %>% as.Date() ) %>% 
        mutate(yq = tsibble::yearquarter(cool_year))


Pop_2050 <- Total_pop_2050 %>%
        as_tsibble(key = c('Country Name', 'Population'),
                            index = 'yq')
Pop_2050$Population = as.numeric(Pop_2050$Population) 
colnames(Pop_2050)[2]<- "iso3c"
Pop_2050 <- Pop_2050 %>%
        mutate(year = year(cool_year)) %>%
        select(`Country Name`, iso3c, Population, year) %>%
        arrange(iso3c, year)

```
###Calculate projection for GDP (PPP) ## 

To find an estimate for predictions of GDP (PPP current international $) to 2050, we use IIASA's SSP scenarios. SSPs estimate growth rate at 5 year intervals out to 2100. We derive a growth rate between the five year increments and providing an annual growth rate for our years of interest (2020-2050). 

```{r}
ssp_scenarios <-read_excel("~/ConstructionProjections/Data/IIASA.xlsx")%>% 
        as.data.table %>%
        select( -Variable, -Model) %>%
        mutate(iso3c = countrycode(Region, "country.name", "iso3c")) %>%
        pivot_longer(cols = starts_with("2"), names_to = "year", values_to = "count") %>%
        mutate(year = as.integer(as.character(year))) %>%
        select(Scenario, iso3c, year, gdp_ppp = count ) %>% 
        arrange(Scenario, iso3c, year) %>%
        mutate(growth_rate = (gdp_ppp / lag(gdp_ppp))^(1/5)) 

ssp_gdp <- CJ(scenario = unique(ssp_scenarios$Scenario), iso3c = unique(ssp_scenarios$iso3c), year = 2020:2050) %>%
        mutate(growth_year = (floor((year - 2020) / 5) + 1) * 5 + 2020,
               gdp_ppp_year = (floor((year - 2020) / 5)) * 5 + 2020) %>%
        left_join(ssp_scenarios %>% select(scenario = Scenario, iso3c, growth_year = year, growth_rate),
                  by = c("scenario", "iso3c", "growth_year")) %>%
        left_join(ssp_scenarios %>% select(scenario = Scenario, iso3c, gdp_ppp_year = year, gdp_ppp_5y = gdp_ppp),
                  by = c("scenario", "iso3c", "gdp_ppp_year")) %>%
        mutate(gdp_ppp = gdp_ppp_5y * (growth_rate) ^ (year - (growth_year - 5)))

head(ssp_gdp)
```
###Take Scenario 2 as Baseline

This provides us with 5 different scenarios for how GDP could change out to 2050. IIASA and UNFCCC recommend using the "middle of road" scenario (Scenario 2) for those looking at a business as usual future world. As such, we take Scenario 2 GDP predictions for our estimates. We also change the unit of observation from billion USD/Yr to base USD/Yr to allow for easier comparison. 

```{r}
ssp_gdp <- ssp_gdp %>%
        filter(scenario == "SSP2 - Review Phase 1") %>%
        mutate(GDP = gdp_ppp * 1000000000)
        
head(ssp_gdp)
```

###check work with a graph for selected countries

To check our data we now select 5 countries at random from different continents and economic backgrounds (USA, Haiti, France, Kenya, and Japan). The figures appear to support general expected trends. 
```{r}
selected_countries <- c("USA", "HTI", "FRA", "KEN", "JPN")

GDP_Filt_2050 <- ssp_gdp %>%
        filter( iso3c %in% selected_countries) %>% 
        select(iso3c, year, GDP) %>% 
        mutate(Country = countrycode(`iso3c`, "iso3c", "country.name")) %>%
        as.data.frame() 

ggplot(GDP_Filt_2050, aes(x = year, y = GDP, color = Country)) +
        geom_line() +
        xlab("Year") + ylab("GDP (PPP)") +
        theme_minimal()

```
###Average Household Size to 2050
We make a decision here not to project household size forward due to lack of data. Instead, we will have the constant household size interact with population to still show change in number of households to allow for change in m2 but without the level of granularity of change in household size. 

###Merge datasets to build figures necessary for floor space m2 to 2050 per country
```{r}
merged_2050 <- Pop_2050%>%
        left_join(average_household, by = c ("iso3c"))%>%
        select(c(`Country Name`, iso3c, Population, year, `Average household size (number of members)`))%>%
        drop_na(Population)

head (merged_2050)
```

### Add average household size for countries not included in UNESDA data set, but included in worldbank population projections 

We find that IIASA data has 193 countries but UNDESA data on household size is missing 31 additional countries. We add these below based on manual search for the figures from national statistics offices of the called countries. 

```{r}
###NA emerging from average household data being incomplete. Add 2.5 for all countries, unless known. Remove non - country regions. 

merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="SWE"] <- 2.15
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="ERI"] <- 5
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="SAU"] <- 5.2
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="ARE"] <- 4.9
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="ISL"] <- 2.5
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="DMA"] <- 3.2
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="LKA"] <- 3.7
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="VIR"] <- 2.14
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="LBY"] <- 5.6
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="LBN"] <- 4.07
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="KWT"] <- 5.5
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="VCT"] <- 3
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="SLB"] <- 5.5
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="SMR"] <- 2.29
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="MNP"] <- 3.26
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="NRU"] <- 6.15
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="FSM"] <- 6
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="MHL"] <- 6.8
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="XKX"] <- 5.5
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="GUM"] <- 3.44
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="GRD"] <- 4
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="GIB"] <- 2.18
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="FRO"] <- 2.85
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="DNK"] <- 2.1
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="CUW"] <- 2.7
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="CHI"] <- 3
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="CPV"] <- 4.98
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="VGB"] <- 2.59
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="BHR"] <- 5.9
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="ATG"] <- 3
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="AND"] <- 2
merged_2050$`Average household size (number of members)`[merged_2050$iso3c =="ASM"] <- 5.6
```

##Add in gdp figures data to merged projection dataset
```{r}
merged_2050_2 <- merged_2050 %>%
        drop_na(`Average household size (number of members)`) %>%
        left_join(ssp_gdp, by = c('iso3c', 'year')) %>%
        select(c(1,2, 3, 4, 5, 11, 12)) %>%
        drop_na('gdp_ppp')
head(merged_2050_2)

rm(merged_2050, undesa_2022, undesa_pd_2022_hh_size_compositions, ssp_gdp, ssp_scenarios, Total_pop_2050, Population, population_projection, GDP_Filt_2050, gdp_per_capita, average_household, Pop_2050, undesa_pd_2022_hh_size_composition)
```

###calculate residential floor space estimate using projections data 

```{r}
residential <- merged_2050_2 %>%
        mutate(average_household_size = as.numeric(`Average household size (number of members)`)) %>%
        mutate(num_households = Population/ average_household_size) %>%
        mutate(gdp_per_capita = GDP / Population) %>%
        mutate(avg_m2_capita = gdp_per_capita / 1000) %>%
        mutate(residential_building_stock_m2 = Population * avg_m2_capita) 

head(residential)
```
### Add regional Weights for total building stock ###

Based on advisory group feedback, add weights for countries on the percent residential vs non residential floor space for each country. Outside of OECD, this number was very difficult to find. We rely on ''Material requirements and impacts of the building sector in the Shared Social Pathways'' by H.M Le Boulzec (2023) from the Journal of Cleaner Production. The estimates provide a residential vs non-residential regional estimate of building floor space between now and 2100 at a five year interval based on SSP 2. 
```{r}
weights <- read_excel("~/ConstructionProjections/Data/weights.xlsx")#Data from Boulzec paper 


BS_2050 <- residential %>%
  select(Country = `Country Name`, iso3c, year, residential_building_stock_m2) %>%
  left_join(read_excel("~/ConstructionProjections/Data/country_region.xlsx") %>% as.data.table, by = "Country") %>%
  mutate(period = ((year - 2020) %/% 5) * 5 + 2020) %>%
  left_join(read_excel("~/ConstructionProjections/Data/weights.xlsx") %>% as.data.table %>%
              select(Region, period = Year, residential), by = c("Region", "period")) %>%
  mutate(total_building_stock_m2 = as.numeric(residential_building_stock_m2) / as.numeric(residential))

```

###Check estimates ###

To check 2050 estimates, we use data from Modelling Global Material Stocks and Flows for Residential and Service Sector buildings to 2050 (Deetman et.al. 2020) which includes estimates on sqm to 2050 for regions of the world. It also includes some further disaggregation for select countries, so we compare estimates for Canada and Japan. We find that our estimates are 23% lower and 10% higher respectively, which is likely due to Deetman et. al. including full building sqm (high-rise and apartments included in their estimates. 

```{r}
Deetman_SQM_Projections <- read_excel("~/ConstructionProjections/Data/Deetman_SQM_Projections.xlsx", 
                                      sheet = "sqmeters_output")
View(Deetman_SQM_Projections)
##Pull for Canada = Region 1  Japan = 23 India = 18 US = 2
select_countries <- (c(1, 2, 18, 23))
Deetman_Select <- Deetman_SQM_Projections %>%
        filter(Region %in% select_countries) %>%
        select(Region, type, area, `2030`, `2050`) 
Sum_row <- Deetman_Select %>%
        group_by(Region) %>%
        summarise(`2030` = sum(`2030`), `2050` = sum(`2050`)) %>%
        mutate(type = "Total")
Deetman_select_Sum <- bind_rows(Deetman_Select, Sum_row) %>%
        mutate(bas_sqm_2030 = `2030` * 1000000) %>%
        mutate(base_sqm_2050 = `2050` * 1000000) %>%
        filter(type == "Total")

### Data in Million sqm need to make in sqm for comparison 
```
Our Estimates: Canada 2030= 2877986259 and 2050 = 3608837481
Our Estimates: Japan 2030 =  7312186774 and 2050 = 8179405377
Our Estimates: India  2030 = 17421199324 and 2050 = 40354419577
Our Estimates: United States 2030 = 34891059816 and 2050 = 45424232511

Deetman Estimates: Canada 2030  = 3243734575 and 2050 = 4080272736
Deetman Estimates: Japan 2030 = 7180523606 and 2050 = 7096909732
Deetman Estimates: India 2030 = 26023014912 and 2050 = 50647169731
Deetman Estimates: US 2030 = 30100197194 and 2050 = 34277734689

For Canada we are 11.9% lower than the Deetman 2030 estimate 12.2% lower than the 2050 estimate
For Japan we are 1.8% higher than Deetman 2030 estimate and 14% higher than the 2050 estimate
For India we are 39% lower than Deetman2030 estimate and 22% lower than the 2050 estimate
For the US we are 14% higher than Deetman 2030 and 27% higher than 2050 estimate 

### Clean and export building stock estimates ###
```{r}
rm(building_stock_m2, Deetman_Select, Deetman_select_Sum, Deetman_SQM_Projections, merged_2050_2, Sum_row, weights)
write_csv(BS_2050, "~/ConstructionProjections/Data/BS_2050.csv")
```