---
title: "Inclusion_WellBeing"
author: "Cassandra Etter"
date: "2024-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Inclusion and Wellbeing Projections

This is the fourth tranche of work which builds on the indicators related to wellbeing and inclusion.

Overview: Data for GlobalABC Project on construction and building indicators and projections
This is the repository containing the source code and documentation for the Global ABC baseline assessment and indicators and projections for gap analysis. 
The point of the project is to 
1) gain a good understanding of the context 
2) identify challenges and opportunities and 
3) measure the objective achievement gap. 

The project was carried out by UNOPS.

###import library
```{r}
library(data.table)
library(dplyr)
library(countrycode)
library(readxl)
library(tidyr)
library(googlesheets4)
library(tidyverse)
```

##2.7 -- Population living in a slum (% of urban Population) (2.7)
Pull Data from UN Habitat on 2020 % of urban population living in slums
Data from UN Habitat available here: https://data.unhabitat.org/pages/housing-slums-and-informal-settlements 

Significant data missing, so apply the regional averages that UNH provide in the same dataset. 
Central Asia and Southern Asia	48.2
Eastern and South-Eastern Asia	21.7
Europe and Northern America	0.7
Latin America and The Carbbean	17.7
Western Asia and Northern Africa	18.7
Oceania	15.3
Sub Saharan Africa	50.2
```{r}
Slum <- read_excel("~/ConstructionProjections/Data/UNH_Slum_Estimates_2000-2020.xlsx") %>%
        mutate(iso3c = countrycode(Country, "country.name", "iso3c")) %>%
        mutate( year = 2020) %>%
        rename(`slum_2.7` = `2020`)
head(Slum)

Well_being_Final <- Slum %>%
        select(Country, iso3c, year, slum_2.7)

```
## Number of people working in the renewable energy sector (5.5)
IRENA's benefits employment data provides a 2022 estimate for all countries on total employment in renewable energy by technology. We take the aggregated ''all technologies'' estimate for our dashboard. 
```{r}
jobs <- read_xlsx("~/ConstructionProjections/Data/IRENA_RE_Jobs_Annual_Review_2023.xlsx") %>%
        filter(Technology == "All technologies") %>%
        mutate(iso3c = countrycode(`Country/area`, "country.name", "iso3c")) %>%
        filter(!is.na(iso3c)) %>%
        rename(`Country` = `Country/area`) 
colnames(jobs)[3] <- "jobs"
jobs <- jobs %>% 
        select(Country, iso3c, Year, jobs) %>%
        mutate(jobs_5.5 = jobs * 1000)
```

###Combine and finalize wellbeing and inclusion 

```{r}
Well_being_Final <- Well_being_Final %>%
        full_join(jobs, by = c("Country","iso3c")) %>%
        select(Country, iso3c, slum_2.7, jobs_5.5)

write_csv(Well_being_Final, "~/ConstructionProjections/Data/Well_being_FINAL.csv")


rm(jobs, Well_being_Final_v2, Slum)
```